---
title: "Analytic Hierarchy Process"
author: "Richard Blackwell"
date: "`r Sys.Date()`"
output: 
  ioslides_presentation:            
    logo: HISW_Logo_RGB_Positive.png
    widescreen: true
    smaller: true
    css: "style.css"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Hierarchy of Criteria

<div style="float: left; width: 20%;">
The diagram shown is an example of a decision tree. There are 3 main criteria (A, B, and C) with A split into 2 sub-criteria (D and E) and C split into 2 sub-criteria (F and G) and E split further into another 2 sub-criteria (H and I)
</div>
<div style="float: right; width: 80%;">
```{r example_hierarchy, echo = FALSE}
DiagrammeR::grViz('ahp001.dot')
```
</div>

## Pairwise Comparisons

For each criteria group (A-B-C), (D-E), (F-G), (H-I) we perform pairwise 
comparisons using the following scale

Criterion **i** compared to criterion **ii**

 - Equally important = 1 
 - Somewhat more important = 3 *(or 1/3)*
 - Definitely more important = 5 *(or 1/5)*
 - Much more important = 7 *(or 1/7)*
 - Very much more important = 9 *(or 1/9)*

If the comparison lies in-between categories use the value between the two categories.

*If criterion **ii** is deemed more important than **i** the reciprocal (1/x) is used as the weighting for that pairwise comparison*

**For example:**
A is somewhere between **much more important** and **very much more important** than B, so we have a value of 8 for the **A vs B** comparison

## Comparison Matrices

For each criteria group we can create a matrix of the comparison

For example:

The matrix of the criteria group (A-B-C) might look like this

| |A|B|C|
|-:|:-:|:-:|:-:|
|A|1|1/4|5|
|B|4|1|1/2|
|C|1/5|2|1|

## Comparison Matrices - Scaling

We then scale each column to equal 1 by dividing each value by the total for that column 

<div style="float: left; width: 40%;">
The following...

| |A|B|C|
|-:|:-:|:-:|:-:|
|A|1|1/4|5|
|B|4|1|1/2|
|C|1/5|2|1|
|Total|5.2|3.25|6.5|
</div>
<div style="float: right; width: 40%;">
... becomes

| |A|B|C|
|-:|:-:|:-:|:-:|
|A|0.19|0.08|0.77|
|B|0.77|0.31|0.08|
|C|0.04|0.62|0.15|
</div>

## Comparison Matrices - Priority Vector

Finally we calculate the mean of each row to give the **priority vector** for the criterion in that row

| |A'|B'|C'|Priority Vector|
|-:|:-:|:-:|:-:|-:|
|A|0.19|0.08|0.77|0.35|
|B|0.77|0.31|0.08|0.38|
|C|0.04|0.62|0.15|0.27|

<p>

Which essentially means that criterion B represents about 38% of the decision, 35% for criterion A and 27% for criterion C.

The same process is undertaken for the sub-criteria to calculate all the priority vectors for each criteria group

## Comparison Matrices - Complete

<div style="float: left; width: 30%;">
Criteria group (D-E)

| |D|E| |D'|E'|Priority Vector|
|-:|:-:|:-:|-|:-:|:-:|-:|
|D|1|1/6||0.14|0.14|0.14|
|E|6|1||0.86|0.86|0.86|
|Total|7|1.17|||||
</div>
<div style="float: left; width: 30%;">
Criteria group (F-G)

| |F|G| |F'|G'|Priority Vector|
|-:|:-:|:-:|-|:-:|:-:|-:|
|F|1|1/3||0.25|0.25|0.25|
|G|3|1||0.75|0.75|0.75|
|Total|4|1.33|||||
</div>
<div style="float: left; width: 30%;">
Criteria group (H-I)

| |H|I| |H'|I'|Priority Vector|
|-:|:-:|:-:|-|:-:|:-:|-:|
|H|1|2||0.67|0.67|0.67|
|I|1/2|1||0.33|0.33|0.33|
|Total|1.5|3|||||
</div>

## Hierarchy of Criteria with Weightings

<div style="float: left; width: 20%;">
The hierarchy diagram can now be updated with the criteria weightings (priority vectors).
</div>
<div style="float: right; width: 80%;">
```{r completed_hierarchy, echo = FALSE}
DiagrammeR::grViz('ahp002.dot')
```
</div>

## Using the Hierarchy to Compare Decisions

We use the same pairwise comparison process as we used to calculate the hierarchy to assess each of the options to make decisions about.

For example, if we have 3 options i, ii, and iii to choose from we compare each option to each other based an each of the criterion that we have selected for the decision making process.

Working up the hierarchy we start by assessing the options in respect to **Criterion I** how does **Option i** compare to **Option ii**, and how does **Option i** compare to **Option ii**, and how does **Option ii** compare to **Option iii**, all in respect to **Criterion I** and we record each of the values in the decision matrix for that criterion.

We then move on to **Criterion H** and do the same, comparing each option for each criterion noted that some of these criteria are calculated from sub-criteri, e.g. E is calculated from H and I, C is calculated from F and G, and A is calculated from D and E.

In the end we have nine 3 x 3 matrices for the pairwaise comparison of each option against each of the criteria

##