---
title: "Analytic Hierarchy Process"
author: "Richard Blackwell"
date: "`r Sys.Date()`"
output: 
  ioslides_presentation:            
    logo: HISW_Logo_RGB_Positive.png
    widescreen: true
    smaller: true
    css: "style.css"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Hierarchy of Criteria

<div style="float: left; width: 20%;">
The diagram shown is an example of a decision tree. 

There are 3 main criteria (A, B, and C) 

Criterion C is formed from 2 sub-criteria (D and E)
</div>
<div style="float: right; width: 80%;">
```{r example_hierarchy, echo = FALSE}
DiagrammeR::grViz('ahp001.dot')
```
</div>

## Pairwise Comparisons

For each criteria group (A-B-C), (D-E) we perform pairwise 
comparisons using the following scale

Criterion **i** compared to criterion **ii**

 - Equally important = 1 
 - Somewhat more important = 3 *(or 1/3)*
 - Definitely more important = 5 *(or 1/5)*
 - Much more important = 7 *(or 1/7)*
 - Very much more important = 9 *(or 1/9)*

If the comparison lies in-between categories use the value between the two categories.

*If criterion **ii** is deemed more important than **i** the reciprocal (1/x) is used as the weighting for that pairwise comparison*

**For example:**
A is somewhere between **much more important** and **very much more important** than B, so we have a value of 8 for the **A vs B** comparison

## Comparison Matrices

For each criteria group we can create a matrix of the comparison

For example:

The matrix of the criteria group (A-B-C) might look like this

| |A|B|C|
|-:|:-:|:-:|:-:|
|A|1|7|3|
|B|1/7|1|1/9|
|C|1/3|9|1|

## Comparison Matrices - Scaling

We then scale each column to equal 1 by dividing each value by the total for that column 

<div style="float: left; width: 40%;">
The following...

| |A|B|C|
|-:|:-:|:-:|:-:|
|A|1|7|3|
|B|1/7|1|1/9|
|C|1/3|9|1|
|Total|1.48|17.00|4.11|
</div>
<div style="float: right; width: 40%;">
... becomes

| |A|B|C|
|-:|:-:|:-:|:-:|
|A|0.68|0.41|0.73|
|B|0.10|0.06|0.03|
|C|0.23|0.53|0.24|
</div>

## Comparison Matrices - Priority Vector

Finally we calculate the mean of each row to give the **priority vector** for the criterion in that row

| |A'|B'|C'|Priority Vector|
|-:|:-:|:-:|:-:|-:|
|A|0.68|0.41|0.73|0.61|
|B|0.10|0.06|0.03|0.06|
|C|0.23|0.53|0.24|0.33|
<p>

Which essentially means that criterion A represents about 61% of the decision, 6% for criterion B and 33% for criterion C.

The same process is undertaken for the sub-criteria to calculate all the priority vectors for each criteria group

## Comparison Matrices - Complete

<div style="float: left; width: 40%;">
Criteria group (D-E)

| |D|E|
|-:|:-:|:-:|
|D|1|3|
|E|1/3|1|
|Total|1.33|4.00|

<p>

| |D'|E'|Priority Vector|
|-:|:-:|:-:|-:|
|D|0.75|0.75|0.75|
|E|0.25|0.25|0.25|

</div>
<div style="float: right; width: 40%;">
All Criteria

| |Priority Vector|
|-:|-:|
|A|0.61|
|B|0.06|
|C|0.33|
|D|0.75|
|E|0.25|

</div>
## Hierarchy of Criteria with Weightings

<div style="float: left; width: 20%;">
The hierarchy diagram can now be updated with the criteria weightings (priority vectors).
</div>
<div style="float: right; width: 80%;">
```{r completed_hierarchy, echo = FALSE}
DiagrammeR::grViz('ahp002.dot')
```
</div>

## Using the Hierarchy to Compare Decisions

We use the same pairwise comparison process as we used to calculate the hierarchy to assess each of the options to make decisions about.

For example, if we have 3 options i, ii, and iii to choose from we compare each option to each other based an each of the criterion that we have selected for the decision making process.

Working up the hierarchy we start by assessing the options in respect to **Criterion E** how does **Option i** compare to **Option ii**, and how does **Option i** compare to **Option ii**, and how does **Option ii** compare to **Option iii**, all in respect to **Criterion E** and we record each of the values in the decision matrix for that criterion.

We then move on to **Criterion D** and do the same, comparing each option for each criterion. 

Note that criterion C is calculated from sub-criteria D and E.

In the end we have four 3 x 3 matrices for the pairwise comparison of each option against each of the criteria

## Option Matrix - Criterion E

In respect to criterion E

- Option i is **the same as** option ii
- Option i is **very much worse than** option iii (or option iii is **very much better than** option i)
- Option ii is **definitely worse than** option iii (or option iii is **definitely better than option** ii)

|E|i|ii|iii|i'|ii'|iii'|PV|
|-:|:-:|:-:|:-:|:-:|:-:|:-:|:-:|
|i|1|1|1/9|0.09|0.14|0.08|0.11|
|ii|1|1|1/5|0.09|0.14|0.15|0.13|
|iii|9|5|1|0.82|0.71|0.76|0.77|
|Total|11.00|7.00|1.31||||

## Option Matrix - Criterion D

In respect to criterion D

- Option i is between **somewhat better** and **definitely better than** option ii
- Option i is **much better than** option iii
- Option ii is **somewhat better than** option iii

|D|i|ii|iii|i'|ii'|iii'|PV|
|-:|:-:|:-:|:-:|:-:|:-:|:-:|:-:|
|i|1|4|7|0.72|0.75|0.64|0.70|
|ii|1/4|1|3|0.18|0.19|0.27|0.21|
|iii|1/7|1/3|1|0.10|0.06|0.09|0.09|
|Total|1.39|5.33|11.00||||

## Option Matrix - Criterion B

In respect to criterion B

- Option i is **much worse than** option ii (or option ii is **much better than** option i)
- Option i is **much worse than** option iii (or option iii is **much better than** option i)
- Option ii is **definitely worse than** option iii (or option iii is **definitely better than** option ii)

Criterion B

|B|i|ii|iii|i'|ii'|iii'|PV|
|-:|:-:|:-:|:-:|:-:|:-:|:-:|:-:|
|i|1|1/7|1/7|0.07|0.02|0.11|0.07|
|ii|7|1|1/5|0.47|0.16|0.15|0.26|
|iii|7|5|1|0.47|0.81|0.74|0.68|
|Total|15.00|6.14|1.34||||

## Option Matrix - Criterion A

In respect to criterion A

- Option i is **somewhat better than** option ii
- Option i is **very much better than** option iii
- Option ii is **much better than** option iii

|A|i|ii|iii|i'|ii'|iii'|PV|
|-:|:-:|:-:|:-:|:-:|:-:|:-:|:-:|
|i|1|3|9|0.69|0.72|0.53|0.65|
|ii|1/3|1|7|0.23|0.24|0.41|0.29|
|iii|1/9|1/7|1|0.08|0.03|0.06|0.06|
|Total|1.44|4.14|17.00||||

## Option Matrix - Criterion C

Criterion C is calculated from the criteria D and E

Criterion C has a PV of 0.33 and D has a PV of 0.75 and E a PV of 0.25

The calculation of the score for option i in respect to criterion C becomes

Option 1 = 0.33 * (score for option i in respect to criterion D + score for option i in respect to criterion E)
Option 1 = 0.33 * ()

## Complete Decision Matrix

|Criterion|Criterion PV|i|ii|iii|
|-:|-:|-:|-:|-:|
|A|0.61|= 0.61 * 0.65<br>= 0.39|= 0.61 * 0.29<br>= 0.18|= 0.61 * 0.06<br>= 0.03|
|B|0.06|= 0.06 * 0.07<br>= 0.00|= 0.06 * 0.26<br>= 0.02|= 0.06 * 0.68<br>= 0.04|
|C|0.33|= 0.33 * (D + E)<br>= 0.33 * (0.53 + 0.03)<br>= 0.18|= 0.33 * (D + E)<br>= 0.33 * (0.16 + 0.03)<br>= 0.06|= 0.33 * (D + E)<br>= 0.33 * (0.06 + 0.19)<br>= 0.08|
|D|0.75|= 0.75 * 0.70<br>= 0.53|= 0.75 * 0.21<br>= 0.16|= 0.75 * 0.09<br>= 0.06|
|E|0.25|= 0.25 * 0.11<br>= 0.03|= 0.25 * 0.13<br>= 0.03|= 0.25 * 0.77<br>= 0.19|
|Goal||0.58|0.26|0.16|

## Hierarchy of Option i Score

<div style="float: left; width: 20%;">
The hierarchy diagram can now be updated for the score for **option i**
</div>
<div style="float: right; width: 80%;">
```{r hierarchy_option_i, echo = FALSE}
DiagrammeR::grViz('ahp003.dot')
```
</div>

## Hierarchy of Option ii Score

<div style="float: left; width: 20%;">
The hierarchy diagram can now be updated for the score for **option ii**
</div>
<div style="float: right; width: 80%;">
```{r hierarchy_option_ii, echo = FALSE}
DiagrammeR::grViz('ahp004.dot')
```
</div>

## Hierarchy of Option iii Score

<div style="float: left; width: 20%;">
The hierarchy diagram can now be updated for the score for **option iii**
</div>
<div style="float: right; width: 80%;">
```{r hierarchy_option_iii, echo = FALSE}
DiagrammeR::grViz('ahp005.dot')
```
</div>

